---
title: "final-data"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(jsonlite)
library(httr)
library(rvest)
library(readxl)
library(tidyverse)
library(mdsr)   
library(tidyr)
library(ggplot2)
library(readr)
library(knitr)
```


### Wrangle and Read Combined Statistical Area Data
``` {r, message=FALSE}

csa_df <- read_csv("../Combined Statistical area/csa-est2017-alldata.csv") %>% 
  select(CSA, CBSA, NAME, LSAD) %>% 
  filter(LSAD == "Micropolitan Statistical Area" | LSAD == "Metropolitan Statistical Area" | LSAD == "Combined Statistical Area") 

csa_title_df <- csa_df %>% 
  filter(LSAD == "Combined Statistical Area") %>% 
  mutate(csa_name = NAME) %>% 
  select(CSA, csa_name) 

titled_csa_df <- inner_join(csa_df, csa_title_df, by = "CSA") %>% 
  filter(LSAD != "Combined Statistical Area") %>% 
  select(CSA, csa_name, CBSA)

colnames(titled_csa_df)[1:3] <- c("csa_id", "csa_name", "msa_id")

# titled_csa_df

```

## Read in Quality of Life Data

### Read in GDP Data
```{r, message=False}

gdp_df <- read_xls("../quality-of-life-data/gdp/gdp-industrytotal.xls", sheet = 1, range = 'A6:S390', col_names=TRUE) %>% 
  gather(key = "Year", value = "gdp", -c("GeoFips", "GeoName")) %>% 
  mutate("GEO.id2" = as.integer(GeoFips), "Year" = as.double(Year)) %>% 
  select(GEO.id2, GeoName, Year, gdp) 

# gdp_df

```

### Read in Latitude Longitude Data
```{r, message=False}

# coordinates for metropolitan areas
metropolitan_coordinates <- read_xlsx("../quality-of-life-data/latitude-longitude/StatisticalAreasLatLong.xlsx", sheet = 1,
                                      range = "A2:Q376", col_names = TRUE)
# coordinates for micropolitan areas
micropolitan_coordinates <- read_xlsx("../quality-of-life-data/latitude-longitude/StatisticalAreasLatLong.xlsx", sheet = 2,
                                      range = "A2:Q583", col_names = TRUE)

# combine metropolitan and micropolitan areas
lat_long_df <- rbind(metropolitan_coordinates, micropolitan_coordinates) %>%
  mutate("GEO.id2" = GEOID, "GEO.display-label" = NAME) %>%
  arrange(NAME) %>%
  select(GEO.id2, `GEO.display-label`, CENTLAT, CENTLON, INTPTLAT, INTPTLON)

# lat_long_df

```

### Read in Population Data
```{r, message=False}

# function to read in population DFs efficiently 
read_population_df <- function(file_name, year) {
  df <- read_csv(file_name) 
  colnames(df)[4] <- "PopEstimate" 
  colnames(df)[5] <- "MOE" 
  df <- df %>% 
    mutate("Year" = year) %>% 
    select(-MOE)
  return(df)
}

# read in population data for years 2007 through 2017
pop_2007 <- read_population_df('../quality-of-life-data/population/2007/ACS_07_1YR_B01003.csv', 2007)
pop_2008 <- read_population_df('../quality-of-life-data/population/2008/ACS_08_1YR_B01003.csv', 2008)
pop_2009 <- read_population_df('../quality-of-life-data/population/2009/ACS_09_1YR_B01003.csv', 2009)
pop_2010 <- read_population_df('../quality-of-life-data/population/2010/ACS_10_1YR_B01003.csv', 2010)
pop_2011 <- read_population_df('../quality-of-life-data/population/2011/ACS_11_1YR_B01003.csv', 2011)
pop_2012 <- read_population_df('../quality-of-life-data/population/2012/ACS_12_1YR_B01003.csv', 2012)
pop_2013 <- read_population_df('../quality-of-life-data/population/2013/ACS_13_1YR_B01003.csv', 2013)
pop_2014 <- read_population_df('../quality-of-life-data/population/2014/ACS_14_1YR_B01003.csv', 2014)
pop_2015 <- read_population_df('../quality-of-life-data/population/2015/ACS_15_1YR_B01003.csv', 2015)
pop_2016 <- read_population_df('../quality-of-life-data/population/2016/ACS_16_1YR_B01003.csv', 2016)
pop_2017 <- read_population_df('../quality-of-life-data/population/2017/ACS_17_1YR_B01003.csv', 2017)

# join all years into one dataframe
population_df <- full_join(pop_2007, pop_2008) %>%
  full_join(pop_2009) %>%
  full_join(pop_2010) %>%
  full_join(pop_2011) %>%
  full_join(pop_2012) %>%
  full_join(pop_2013) %>%
  full_join(pop_2014) %>%
  full_join(pop_2015) %>%
  full_join(pop_2016) %>%
  full_join(pop_2017)

# population_df

```

### Read in Census Quality of Life Data
```{r, message=FALSE}

# function to clean up data frames from 2007-2009
# (the census changed variable names after 2009)
clean_data_2007_2009 <- function(file_name, year) {
  df <- read_csv(file_name) %>% 
    mutate('Median household income (dollars)' = HC01_EST_VC69,
           'Percent workers commuting by public transportation' = HC02_EST_VC23,
           'Percent population unemployed' = HC02_EST_VC06,
           'Year' = year) %>% 
     select(Year, GEO.id, GEO.id2, `GEO.display-label`, 
           `Median household income (dollars)`, 
           `Percent workers commuting by public transportation`,
           `Percent population unemployed`)
  return(df)
}

# function to clean up data frames from 2010-2017
clean_data_2010_2017 <- function(file_name, year) {
  df <- read_csv(file_name) %>% 
    mutate('Median household income (dollars)' = HC01_VC85,
           'Percent workers commuting by public transportation' = HC03_VC31,
           'Percent population unemployed' = HC03_VC08,
           'Percent no health insurance coverage' = HC03_VC134,
          
            # for all people 18 years and older:
           'Percent income below poverty level' = as.numeric(HC03_VC171),
           'Year' = year ) %>% 
     select(Year, GEO.id, GEO.id2, `GEO.display-label`, 
           `Median household income (dollars)`, 
           `Percent workers commuting by public transportation`,
           `Percent population unemployed`,
           `Percent no health insurance coverage`,
           `Percent income below poverty level`)
            
  return(df)
}

# read in census data for  years 2007 through 2017
census_data_2007 <- clean_data_2007_2009("../quality-of-life-data/census-data/2007/ACS_07_1YR_DP3.csv", 2007)
census_data_2008 <- clean_data_2007_2009("../quality-of-life-data/census-data/2008/ACS_08_1YR_DP3.csv", 2008)
census_data_2009 <- clean_data_2007_2009("../quality-of-life-data/census-data/2009/ACS_09_1YR_DP3.csv", 2009)
census_data_2010 <- clean_data_2010_2017("../quality-of-life-data/census-data/2010/ACS_10_1YR_DP03.csv", 2010)
census_data_2011 <- clean_data_2010_2017("../quality-of-life-data/census-data/2011/ACS_11_1YR_DP03.csv", 2011)
census_data_2012 <- clean_data_2010_2017("../quality-of-life-data/census-data/2012/ACS_12_1YR_DP03.csv", 2012)
census_data_2013 <- clean_data_2010_2017("../quality-of-life-data/census-data/2013/ACS_13_1YR_DP03.csv", 2013)
census_data_2014 <- clean_data_2010_2017("../quality-of-life-data/census-data/2014/ACS_14_1YR_DP03.csv", 2014)
census_data_2015 <- clean_data_2010_2017("../quality-of-life-data/census-data/2015/ACS_15_1YR_DP03.csv", 2015)
census_data_2016 <- clean_data_2010_2017("../quality-of-life-data/census-data/2016/ACS_16_1YR_DP03.csv", 2016)
census_data_2017 <- clean_data_2010_2017("../quality-of-life-data/census-data/2017/ACS_17_1YR_DP03.csv", 2017)

# join all years into one dataframe
census_qol_df <- full_join(census_data_2007, census_data_2008) %>% 
  full_join(census_data_2009) %>% 
  full_join(census_data_2010) %>% 
  full_join(census_data_2011) %>% 
  full_join(census_data_2012) %>% 
  full_join(census_data_2013) %>% 
  full_join(census_data_2014) %>% 
  full_join(census_data_2015) %>% 
  full_join(census_data_2016) %>% 
  full_join(census_data_2017) 

  ## If we want to fix GEO.display-label column then need code similar to this:
  # %>% 
  # separate(col = `GEO.display-label`, into = c("City","State","Area_type"), 
  #         sep = c(","), remove = TRUE)


# census_qol_df

```

### Join All Quality of Life Dataframes
```{r, message=FALSE}

# join all quality of life dataframes
qol_df <- left_join(population_df, census_qol_df, by = c("Year", "GEO.id", "GEO.id2", "GEO.display-label")) %>% 
  left_join(gdp_df, by = c("GEO.id2", "Year")) %>% 
  left_join(lat_long_df, by = c("GEO.id2", "GEO.display-label")) %>% 
  select(-c(GEO.id, GeoName))

# qol_df

```
## Read in Transit Data

### Functions to Clean up FTA Dataframes
```{r, message=FALSE}

#function to clean the modes of transit
clean_modes <- function(df) {
  rail_abbr <- c('HR', 'LR', 'SR', 'CR', 'IP', 'MG', 'YR', 'AR', 'CC')
  nr_bus_abbr <- c('MB', 'TB', 'CB', 'RB', 'PB', 'JT')
  nr_other_abbr <- c('VP', 'DR', 'DT', 'TR', 'FB')
  
  mutate(df, modes_clean = 
      case_when(
        Modes %in% rail_abbr ~ "Rail",
        Modes %in% nr_bus_abbr ~ "Non-Rail Bus",
        Modes %in% nr_other_abbr ~ "Non-Rail Other",
        TRUE ~ "OTHER"))  %>% 
      filter(modes_clean != "OTHER")
  
}


# function to read in and clean up FTA dataframes
clean_fta_df <- function(sheet_number, col_name) {
  
  df <- read_xlsx("../FTA-data/FTA-dataframes/FTA_September_2018.xlsx", sheet=sheet_number, range="A1:HB2129", col_names=TRUE) 
  
  df <- df %>% 
    filter(`Reporter Type` == "Full Reporter") %>% 
    filter(Active == "Active") %>% 
     gather(key = "Date", 
          value = Name_of_Var, 
           -`5 digit NTD ID`, 
           -`4 digit NTD ID`, 
           -Agency, 
           -Active, 
          -`Reporter Type`, 
          -UZA, 
          -`UZA Name`, 
          -Modes, 
          -TOS)  %>% 
    clean_modes()

    df[,col_name] <- df[,'Name_of_Var']
    df[,'Name_of_Var'] <- NULL

    return(df)
}

# function to clean up FTA funding dataframes
clean_funding_df <- function(sheet_number, col_name) {
  df <- read_xlsx("../FTA-data/FTA-dataframes/TS1.1TimeSeriesOpCapFundingSummary_4 (2).xlsx", sheet=sheet_number, range="A1:AO2943", col_names=TRUE)
  
  df <- df %>% 
    filter(`Reporter Type` == "Full Reporter") %>% 
      select(-`Reporter Type`) %>%
      gather(key = "Year",
           value = Name_of_Var,
           -`Last Report Year`,
           -`NTD ID`,
            -`Legacy NTD ID`,
            -`Agency Name`,
           -`Agency Status`,
           -City,
           -State,
           -`Census Year`,
           -`Primary UZA Name`,
           -UZA,
           -`UZA Area SQ Miles`,
           -`UZA Population`,
           -`2017 Status`)

    df[,col_name] <- df[,'Name_of_Var']
    df[,'Name_of_Var'] <- NULL

    return(df)
}

```

### Read in Master FTA Dataframe
```{r, message=FALSE}

# read in DF with master information about each transportation agency 
# and the relevant Urbanized Area (UZA)
master_fta_df <- read_xlsx("../FTA-data/FTA-dataframes/FTA_September_2018.xlsx", sheet=2, range="A1:Y2129", col_names=TRUE) %>% 
  filter(`Reporter Type` == "Full Reporter") %>% 
  filter(Active == "Active") 

# master_fta_df

```

### Read in UPT, VRM, & VRH from Master FTA Dataframe
```{r, message=FALSE}

# read in Unlinked Passenger Trips (UPTs) dataframe
upt_df<- clean_fta_df(3, "UPTs") %>% 
  mutate(UPTs = as.numeric(UPTs))%>% 
  mutate(Year = substr(Date, 4, 5)) %>% 
  group_by(UZA, Year, modes_clean)%>% 
  mutate(UPTs = sum(UPTs, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, UPTs) %>% 
  unique()


# read in Vehicle Revenue Miles (VRMs) dataframe
vrm_df<- clean_fta_df(4, "VRMs") %>%  
  mutate(VRMs = as.numeric(VRMs)) %>% #convert UPT from characters into doubles
  mutate(Year = substr(Date, 4, 5)) %>% 
  group_by(UZA, Year, modes_clean)%>% 
  mutate(VRMs = sum(VRMs, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, VRMs ) %>% 
  unique()

# read in Vehicle Revenue Hours (VRHs) dataframe
vrh_df<- clean_fta_df(5, "VRHs") %>% 
  mutate(VRHs = as.numeric(VRHs)) %>% #convert UPT from characters into doubles
  mutate(Year = substr(Date, 4, 5)) %>% 
  group_by(UZA, Year, modes_clean)%>% 
  mutate(VRHs = sum(VRHs, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, VRHs) %>% 
  unique()

```

### Read in Transit Stations Dataframe
```{r, message=FALSE}

# read in transit stations dataframe
transit_stations_df1 <- read_xlsx("../FTA-data/FTA-dataframes/Transit Stations_0.xlsx", sheet=1, range="A1:L1100", col_names=TRUE) %>%
  filter(`Reporter Type` == "Full Reporter") %>% 
  mutate(Modes = Mode) %>% 
  select(`NTD ID`, `Agency Name`, Modes, `Total Stations`) %>% 
  clean_modes() %>% 
  select(-Modes)
  
# join transit stations dataframe with master FTA dataframe
# (note that this potentially double counts transit stations by used by multiple transit networks)
transit_stations_df <- left_join(master_fta_df, transit_stations_df1, by = c("5 digit NTD ID" = "NTD ID" )) %>% 
  select(Agency, UZA, `UZA Name`,`Agency Name`, modes_clean, `Total Stations`) %>% 
  group_by(UZA, modes_clean) %>% 
  mutate(Total_Stations_2017 = sum(`Total Stations`, na.rm = TRUE)) %>% 
  select(UZA, modes_clean,Total_Stations_2017) %>% 
  filter(!is.na(modes_clean)) %>% 
  unique() 


```

### Read in Transit Funding Data
```{r, message=FALSE}

# read in Total Funding dataframe
Total_Funding_df <- clean_funding_df(3, "Total_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(Total_Funding = sum(Total_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, Total_Funding) %>% 
  unique()

# read in Federal Funding dataframe
Federal_Funding_df <- clean_funding_df(4, "Federal_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(Federal_Funding = sum(Federal_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, Federal_Funding) %>% 
  unique()

# read in State Funding dataframe
State_Funding_df <- clean_funding_df(5, "State_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(State_Funding = sum(State_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, State_Funding) %>% 
  unique()

# read in Local Funding dataframe
Local_Funding_df <- clean_funding_df(6, "Local_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(Local_Funding = sum(Local_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, Local_Funding) %>% 
  unique()

# read in Other Funding dataframe (aka Fares and other assorted revenue) 
Other_Funding_df <- clean_funding_df(7, "Other_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(Other_Funding = sum(Other_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, Other_Funding) %>% 
  unique()
```

### Join FTA Funding Dataframes
```{r, message=FALSE}

# join all Funding dataframes
#NOTE: The sum of federal, state, local and other funding do not equal total funding. Something to note and potentially figure out at some point
fta_funding_df <- full_join(Total_Funding_df, Federal_Funding_df, by = c("UZA", "Year")) %>% 
  full_join(State_Funding_df, by = c("UZA", "Year")) %>% 
  full_join(Local_Funding_df, by = c("UZA", "Year")) %>% 
  full_join(Other_Funding_df, by = c("UZA", "Year")) 

# fta_funding_df

```

### Join All FTA Dataframes
```{r, message=FALSE}

final_master_fta_df <- master_fta_df %>% 
  mutate( UZA_Pop_2010 = `UZA Population`) %>% 
  select(UZA, `UZA Area SQ Miles`, UZA_Pop_2010)

# join master FTA dataframe with UPT, VRM, VRH
fta_df <- full_join(final_master_fta_df, upt_df, by= c("UZA")) %>% 
  full_join(vrm_df, by= c("UZA", "UZA Name", "Year", "modes_clean")) %>% 
  full_join(vrh_df, by= c("UZA", "UZA Name", "Year", "modes_clean")) %>% 
  mutate(temp = 20) %>% 
  unite(Year, c(temp, Year), sep = "", remove = TRUE) %>% 
  unique() %>% 
  # join transit stations dataframes
  # (potentially double counts transit stations that are used by multiple transit networks)
  full_join(transit_stations_df, by = c("UZA", "modes_clean")) %>% 
  # joining FTA funding data
  full_join(fta_funding_df, by=c("UZA", "Year")) %>%  
    filter(Year >= 2001)

# fta_df

```

### Add Identifiers to FTA Dataframe
``` {r, message=FALSE}

# get census key dataframe
census_UZA_MSA_key <- read_csv("../census-FTA-joining-key/UZA_MSA_join.txt") %>% 
  mutate("GEO.id2" = CBSA) %>% 
  arrange(desc(UAPOP)) %>% 
  select(GEO.id2, UA, UANAME, UAPOP)

# join FTA dataframes here
final_fta_df <- left_join(census_UZA_MSA_key, fta_df,
                    by = c("UAPOP" = "UZA_Pop_2010")) %>% 
  mutate(Year = as.numeric(Year))%>% 
  filter(Year > 2006, Year < 2018) %>% 
  filter(!is.na(`UZA Name`)) %>% 
  filter(UANAME != "Not in a 2010 urban area" ) %>% 
  filter(UAPOP>100000) %>% 
  select(-UANAME, -UZA)

final_fta_df_csa <- inner_join(titled_csa_df, final_fta_df, by = c("msa_id"= "GEO.id2")) %>% 
  group_by(csa_id) %>% 
  mutate(`UZA Area SQ Miles` = sum(`UZA Area SQ Miles`)) %>%
  group_by(Year, modes_clean) %>% 
  mutate(UPTs = sum(UPTs)) %>%
  mutate(VRMs = sum(VRMs)) %>%
  mutate(VRHs = sum(VRHs)) 
  
```



## Join Quality of Life and Transit Data

```{r, message=FALSE}

# join final quality of life and fta datasets and filter out non-urban areas 
final_df <- left_join(qol_df, final_fta_df, by=c("GEO.id2", "Year"))%>% 
  filter(UA != 99999) %>% 
  mutate(gdp = gdp*1000000) 
 
# clarify variables names
  colnames(final_df)[1:28] <- c("msa_id", 
                               "msa_name", 
                               "pop_estimate_msa", 
                               "year", 
                               "median_household_income_msa",
                               "percent_workers_commuting_by_public_transit_msa",
                               "percent_unemployed_msa",
                               "percent_no_insurance_msa",
                               "percent_below_poverty_level",
                               "gdp_msa",
                               "centlat",
                               "centlon",
                               "intptlat",
                               "intptlon",
                               "ua_census_id",
                               "ua_pop_2010",
                               "ua_sq_miles_2010",
                               "ua_fta_name",
                               "transit_modes",
                               "passenger_trips",
                               "vehicle_miles",
                               "vehicle_hours",
                               "total_stations_2017",
                               "total_funding",
                               "federal_funding",
                               "state_funding",
                               "local_funding",
                               "other_funding")
  
  
  
```

## Adding Per Capita Variables 
```{r, message = FALSE}

# function to scale from 0 to 1
scale_0_1 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

#Per Capita Transit Variables
per_capita_df <- final_df %>% 
  mutate(per_capita_gdp =  gdp_msa/pop_estimate_msa)

scaled_df <- per_capita_df %>% 
  mutate(scaled_tot = scale_0_1(per_capita_total_funding)) %>% 
  mutate(scaled_fed = scale_0_1(per_capita_federal_funding)) %>%  
  mutate(scaled_sta = scale_0_1(per_capita_state_funding)) %>% 
  mutate(scaled_loc = scale_0_1(per_capita_local_funding)) %>%  
  mutate(scaled_oth = scale_0_1(per_capita_other_funding)) %>% 
  mutate(scaled_trips = scale_0_1(per_capita_passenger_trips)) %>% 
  mutate(scaled_miles = scale_0_1(per_capita_vehicle_miles)) %>% 
  mutate(scaled_hours = scale_0_1(per_capita_vehicle_hours)) %>% 
  mutate(avg_scaled = round(((scaled_tot + scaled_trips + scaled_miles +scaled_hours)/4), 4)) 

  scaled_df$avg_scaled <- round(scaled_df$avg_scaled, 4)
  
 
  

```

## Head(df) & Write DF 
```{r, message = FALSE}

head(final_df)

write.csv(final_df, "FINAL_DATA.csv", row.names=F)

saveRDS(final_df, file = "final_data.RDA")

readRDS("../final-data/final_data.RDA")
```