---
title: "STAT-231 Final Project: Technical Report"
author: "Martin Glusker, Nicole Frontero, Oliver Baldwin-Edwards"
date: 'Due: 12/18/18'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(httr)
library(rvest)
library(readxl)
library(tidyverse)
library(mdsr)   
library(tidyr)
library(ggplot2)
library(readr)
library(knitr)
library(roxygen2)#needed for gluskr documentation
library(rpart)
library(gluskr)
options(warn=-1)
```

# Abstract:


# Introduction:


# Data:


## Quality of Life Data
### Read in GDP Data
This creates a data frame with gdp for each metropolitan and micropolitan statistical area from the U.S. census from years 2001-2017.
```{r, messages=FALSE}

gdp_df <- read_xls("../quality-of-life-data/gdp/gdp-industrytotal.xls", 
                   sheet = 1, range = 'A6:S390', col_names=TRUE) %>% 
  gather(key = "Year", value = "gdp", -c("GeoFips", "GeoName")) %>% 
  mutate("GEO.id2" = as.integer(GeoFips), "Year" = as.double(Year)) %>% 
  select(GEO.id2, GeoName, Year, gdp) 

```

### Read in Latitude Longitude Data
This creates a data frame with latitudinal and longitudinal coordinates for each metropolitan and micropolitan statistical area. 
```{r, messages=FALSE}

# coordinates for metropolitan areas
metropolitan_coordinates <- read_xlsx(
  "../quality-of-life-data/latitude-longitude/StatisticalAreasLatLong.xlsx",
  sheet = 1, range = "A2:Q376", col_names = TRUE)

# coordinates for micropolitan areas
micropolitan_coordinates <- read_xlsx(
  "../quality-of-life-data/latitude-longitude/StatisticalAreasLatLong.xlsx",
  sheet = 2, range = "A2:Q583", col_names = TRUE)

# combine metropolitan and micropolitan areas
lat_long_df <- rbind(metropolitan_coordinates, micropolitan_coordinates) %>%
  mutate("GEO.id2" = GEOID, "GEO.display-label" = NAME) %>%
  arrange(NAME) %>%
  select(GEO.id2, `GEO.display-label`, CENTLAT, CENTLON, INTPTLAT, INTPTLON)

```

### Read in Population Data
This code chunk reads in the population estimates from the U.S. Census Bureau for each year between 2007 and 2017 and joins each year's data into one data frame.
```{r, messages=FALSE, warning=FALSE}

# function to read in population DFs efficiently 
read_population_df <- function(file_name, year) {
  
  df <- read_csv(file_name) 
  
  colnames(df)[4] <- "PopEstimate" 
  colnames(df)[5] <- "MOE" 
  df <- df %>% 
    mutate("Year" = year) %>% 
    select(-MOE)
  
  return(df)
}

# read in population data for years 2007 through 2017
pop_2007 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2007/ACS_07_1YR_B01003.csv', 2007))
pop_2008 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2008/ACS_08_1YR_B01003.csv', 2008))
pop_2009 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2009/ACS_09_1YR_B01003.csv', 2009))
pop_2010 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2010/ACS_10_1YR_B01003.csv', 2010))
pop_2011 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2011/ACS_11_1YR_B01003.csv', 2011))
pop_2012 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2012/ACS_12_1YR_B01003.csv', 2012))
pop_2013 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2013/ACS_13_1YR_B01003.csv', 2013))
pop_2014 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2014/ACS_14_1YR_B01003.csv', 2014))
pop_2015 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2015/ACS_15_1YR_B01003.csv', 2015))
pop_2016 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2016/ACS_16_1YR_B01003.csv', 2016))
pop_2017 <- suppressMessages(read_population_df(
  '../quality-of-life-data/population/2017/ACS_17_1YR_B01003.csv', 2017))


# join all years into one dataframe
suppressMessages(
population_df <- full_join(pop_2007, pop_2008) %>%
  full_join(pop_2009) %>%
  full_join(pop_2010) %>%
  full_join(pop_2011) %>%
  full_join(pop_2012) %>%
  full_join(pop_2013) %>%
  full_join(pop_2014) %>%
  full_join(pop_2015) %>%
  full_join(pop_2016) %>%
  full_join(pop_2017))

```

### Read in Census Quality of Life Data
This data frame consists of quality of life related variables obtained from the U.S. Census Bureau from 2007-2017. The Census Bureau changed the names and descriptions of their variables after 2009 which thus meant two different functions were needed to accurately read in the desired variables. There were also variables introduced in 2010 that we wanted to include in our data frame (`percent_no_insurance_msa` and `percent_below_poverty_level`). Because they didn't exist until 2010 they are recorded as NAs from 2007-2009.  
```{r, messages=FALSE, warnings=FALSE}

# function to clean up data frames from 2007-2009
# (the census changed variable names after 2009)
clean_data_2007_2009 <- function(file_name, year) {
  df <- read_csv(file_name) %>% 
    mutate('Median household income (dollars)' = HC01_EST_VC69,
           'Percent workers commuting by public transportation' = HC02_EST_VC23,
           'Percent population unemployed' = HC02_EST_VC06,
           'Year' = year) %>% 
     select(Year, GEO.id, GEO.id2, `GEO.display-label`, 
           `Median household income (dollars)`, 
           `Percent workers commuting by public transportation`,
           `Percent population unemployed`)
  return(df)
}

# function to clean up data frames from 2010-2017
clean_data_2010_2017 <- function(file_name, year) {
  df <- read_csv(file_name) %>% 
    mutate('Median household income (dollars)' = HC01_VC85,
           'Percent workers commuting by public transportation' = HC03_VC31,
           'Percent population unemployed' = HC03_VC08,
           'Percent no health insurance coverage' = HC03_VC134,
          
            # for all people 18 years and older:
           'Percent income below poverty level' = as.numeric(HC03_VC171),
           'Year' = year ) %>% 
     select(Year, GEO.id, GEO.id2, `GEO.display-label`, 
           `Median household income (dollars)`, 
           `Percent workers commuting by public transportation`,
           `Percent population unemployed`,
           `Percent no health insurance coverage`,
           `Percent income below poverty level`)
            
  return(df)
}

# read in census data for  years 2007 through 2017
census_data_2007 <- suppressMessages(clean_data_2007_2009(
  "../quality-of-life-data/census-data/2007/ACS_07_1YR_DP3.csv", 2007))
census_data_2008 <- suppressMessages(clean_data_2007_2009(
  "../quality-of-life-data/census-data/2008/ACS_08_1YR_DP3.csv", 2008))
census_data_2009 <- suppressMessages(clean_data_2007_2009(
  "../quality-of-life-data/census-data/2009/ACS_09_1YR_DP3.csv", 2009))
census_data_2010 <- suppressMessages(clean_data_2010_2017(
  "../quality-of-life-data/census-data/2010/ACS_10_1YR_DP03.csv", 2010))
census_data_2011 <- suppressMessages(clean_data_2010_2017(
  "../quality-of-life-data/census-data/2011/ACS_11_1YR_DP03.csv", 2011))
census_data_2012 <- suppressMessages(clean_data_2010_2017(
  "../quality-of-life-data/census-data/2012/ACS_12_1YR_DP03.csv", 2012))
census_data_2013 <- suppressMessages(clean_data_2010_2017(
  "../quality-of-life-data/census-data/2013/ACS_13_1YR_DP03.csv", 2013))
census_data_2014 <- suppressMessages(clean_data_2010_2017(
  "../quality-of-life-data/census-data/2014/ACS_14_1YR_DP03.csv", 2014))
census_data_2015 <- suppressMessages(clean_data_2010_2017(
  "../quality-of-life-data/census-data/2015/ACS_15_1YR_DP03.csv", 2015))
census_data_2016 <- suppressMessages(clean_data_2010_2017(
  "../quality-of-life-data/census-data/2016/ACS_16_1YR_DP03.csv", 2016))
census_data_2017 <- suppressMessages(clean_data_2010_2017(
  "../quality-of-life-data/census-data/2017/ACS_17_1YR_DP03.csv", 2017))

# join all years into one dataframe
census_qol_df <- suppressMessages(full_join(census_data_2007, census_data_2008) %>% 
  full_join(census_data_2009) %>% 
  full_join(census_data_2010) %>% 
  full_join(census_data_2011) %>% 
  full_join(census_data_2012) %>% 
  full_join(census_data_2013) %>% 
  full_join(census_data_2014) %>% 
  full_join(census_data_2015) %>% 
  full_join(census_data_2016) %>% 
  full_join(census_data_2017) )

```

### Join All Quality of Life Dataframes
This joins all of the quality of life related data into one data frame.  
```{r, messages=FALSE}

# join all quality of life dataframes
qol_df <- left_join(population_df, census_qol_df, 
  by = c("Year", "GEO.id", "GEO.id2", "GEO.display-label")) %>% 
  left_join(gdp_df, by = c("GEO.id2", "Year")) %>% 
  left_join(lat_long_df, by = c("GEO.id2", "GEO.display-label")) %>% 
  select(-c(GEO.id, GeoName))

```

## Transit Data
### Functions to Clean up FTA Dataframes
These are three functions used to wrangle the transit data from the FTA. `clean_modes()` takes in a data frame with variables `Modes` and groups categories them into a new variable called `modes_clean`. `clean_fta_df()` is a function to clean FTA data frames. `clean_funding_df()` is a function to clean the FTA's transit agency funding data. 
```{r, messages=FALSE}

#function to clean the modes of transit
clean_modes <- function(df) {
  rail_abbr <- c('HR', 'LR', 'SR', 'CR', 'IP', 'MG', 'YR', 'AR', 'CC')
  bus_abbr <- c('MB', 'TB', 'CB', 'RB', 'PB', 'JT')
  other_abbr <- c('VP', 'DR', 'DT', 'TR', 'FB', 'OT', 'OR')
  
  mutate(df, modes_clean = 
      case_when(
        Modes %in% rail_abbr ~ "Rail",
        Modes %in% bus_abbr ~ "Bus",
        Modes %in% other_abbr ~ "Other",
        TRUE ~ "OTHER"))  %>% 
      filter(modes_clean != "OTHER")
  
}

# function to read in and clean up FTA dataframes
clean_fta_df <- function(sheet_number, col_name) {
  
  df <- read_xlsx("../FTA-data/FTA-dataframes/TS2.1TimeSeriesOpExpSvcModeTOS_2.xlsx", 
                  sheet=sheet_number, range="A1:AR5659", col_names=TRUE) 
  
  df <- df %>% 
    filter(`Reporter Type` == "Full Reporter") %>%
     gather(key = "Year", 
          value = Name_of_Var, 
           -`Last Report Year`,
           -`NTD ID`, 
           -`Legacy NTD ID`, 
           -`Agency Name`, 
           -`Agency Status`, 
           -`Reporter Type`, 
           -City, 
           -State, 
           -`Census Year`,
           -`UZA Name`,
           -UZA,
           -`UZA Area SQ Miles`,
           -`UZA Population`,
           -`2017 Status`,
          -Mode,
          -Service, 
          -`Mode Status`)  %>% 
    mutate(Modes = Mode) %>% 
    clean_modes() %>% 
      select(-Mode) %>% 
    filter(Year >= 2007) %>% 
    filter(UZA !=0)

    df[,col_name] <- df[,'Name_of_Var']
    df[,'Name_of_Var'] <- NULL

    return(df)
}

# function to clean up FTA funding dataframes
clean_funding_df <- function(sheet_number, col_name) {
  df <- read_xlsx(
    "../FTA-data/FTA-dataframes/TS1.1TimeSeriesOpCapFundingSummary_4 (2).xlsx", 
                  sheet=sheet_number, range="A1:AO2943", col_names=TRUE)
  
  df <- df %>% 
    filter(`Reporter Type` == "Full Reporter") %>% 
      select(-`Reporter Type`) %>%
      gather(key = "Year",
           value = Name_of_Var,
           -`Last Report Year`,
           -`NTD ID`,
            -`Legacy NTD ID`,
            -`Agency Name`,
           -`Agency Status`,
           -City,
           -State,
           -`Census Year`,
           -`Primary UZA Name`,
           -UZA,
           -`UZA Area SQ Miles`,
           -`UZA Population`,
           -`2017 Status`)

    df[,col_name] <- df[,'Name_of_Var']
    df[,'Name_of_Var'] <- NULL

    return(df)
}

```

### Main Transit Variables
Creates and tidies data frame for each transit variable (UPT, VRM, VRH, DRM, PMT, expenses, and fares).
```{r, messages=FALSE}

expenses_fta_df <- clean_fta_df(3, "total_expenses") %>% 
  group_by(UZA, Year, modes_clean) %>% 
  mutate(total_expenses = sum(total_expenses, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, total_expenses) %>% 
  unique()

fares_fta_df <- clean_fta_df(8, "total_fares") %>% 
  group_by(UZA, Year, modes_clean) %>% 
  mutate(total_fares = sum(total_fares, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, total_fares) %>% 
  unique()

drm_fta_df <- clean_fta_df(9, "drm") %>% 
  group_by(UZA, Year, modes_clean) %>% 
  mutate(drm = sum(drm, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, drm) %>% 
  unique()

vrm_fta_df <- clean_fta_df(11, "vrm") %>% 
  group_by(UZA, Year, modes_clean) %>% 
  mutate(vrm = sum(vrm, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, vrm) %>% 
  unique()

vrh_fta_df <- clean_fta_df(12, "vrh") %>% 
  group_by(UZA, Year, modes_clean) %>% 
  mutate(vrh = sum(vrh, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, vrh) %>% 
  unique()

upt_fta_df <- clean_fta_df(13, "upt") %>% 
  group_by(UZA, Year, modes_clean) %>% 
  mutate(upt = sum(upt, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, upt) %>% 
  unique()

pmt_fta_df <- clean_fta_df(14, "pmt") %>% 
  group_by(UZA, Year, modes_clean) %>% 
  mutate(pmt = sum(pmt, na.rm = TRUE)) %>% 
  select(UZA, `UZA Name`, Year, modes_clean, pmt) %>% 
  unique()

```

### Join FTA Variable Dataframes
Joins above transit data frames into one data frame.
```{r, messages=FALSE}

# create file with master info
master_fta_df <- clean_fta_df(3, "total_expenses") %>% 
  select(UZA, `UZA Area SQ Miles`, `UZA Population`)

# join all FTA Variable dataframes
primary_variable_fta_df <- 
  inner_join(master_fta_df, expenses_fta_df, by= c("UZA")) %>% 
  inner_join(fares_fta_df,    by= c("UZA", "UZA Name", "Year", "modes_clean")) %>% 
  inner_join(drm_fta_df,      by= c("UZA", "UZA Name", "Year", "modes_clean")) %>% 
  inner_join(vrh_fta_df,      by= c("UZA", "UZA Name", "Year", "modes_clean")) %>% 
  inner_join(vrm_fta_df,      by= c("UZA", "UZA Name", "Year", "modes_clean")) %>% 
  inner_join(pmt_fta_df,      by= c("UZA", "UZA Name", "Year", "modes_clean")) %>%
  inner_join(upt_fta_df,      by= c("UZA", "UZA Name", "Year", "modes_clean")) %>% 
  unique()

```

### Read in Transit Stations Dataframe
This chunk reads in the transit stations data frame from the FTA and cleans it. It then joins the master data frame to the transit stations data frame.
```{r, messages=FALSE}

# Read in Master FTA Dataframe to use 
# as joining key from transit agency to UZA
transit_stations_master_fta_df <- read_xlsx(
  "../FTA-data/FTA-dataframes/FTA_September_2018.xlsx",
  sheet=2, range="A1:Y2129", col_names=TRUE) %>% 
  filter(`Reporter Type` == "Full Reporter") %>%  
  select(UZA, `5 digit NTD ID`)

# read in transit stations dataframe
transit_stations_df <- read_xlsx(
  "../FTA-data/FTA-dataframes/Transit Stations_0.xlsx",
  sheet=1, range="A1:L1100", col_names=TRUE) %>%
  filter(`Reporter Type` == "Full Reporter") %>% 
  mutate(Modes = Mode) %>% 
  clean_modes() %>% 
  select(-Mode) %>% 
  select(`NTD ID`, modes_clean, `Total Stations`)
  
  
# Creates final transit stations df by IDing with UZA ids 
# and joining those ids onto the stations data via Agnecy ID
# NOTE: this potentially double counts transit stations 
# that are used by multiple transit networks
final_transit_stations_df <- left_join(transit_stations_master_fta_df, 
  transit_stations_df, by = c("5 digit NTD ID" = "NTD ID" )) %>% 
  group_by(UZA, modes_clean) %>% 
  mutate(Total_Stations_2017 = sum(`Total Stations`, na.rm = TRUE)) %>% 
  select(UZA, modes_clean, Total_Stations_2017) %>% 
  filter(!is.na(modes_clean)) %>% 
  unique() 

```

### Read in Transit Funding Data
Creates and tidies data frame for each transit funding variable (total funding, state funding, local funding, other funding).
```{r, messages=FALSE}

# read in Total Funding data frame
Total_Funding_df <- clean_funding_df(3, "Total_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(Total_Funding = sum(Total_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, Total_Funding) %>% 
  unique()

# read in Federal Funding data frame
Federal_Funding_df <- clean_funding_df(4, "Federal_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(Federal_Funding = sum(Federal_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, Federal_Funding) %>% 
  unique()

# read in State Funding data frame
State_Funding_df <- clean_funding_df(5, "State_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(State_Funding = sum(State_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, State_Funding) %>% 
  unique()

# read in Local Funding data frame
Local_Funding_df <- clean_funding_df(6, "Local_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(Local_Funding = sum(Local_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, Local_Funding) %>% 
  unique()

# read in Other Funding data frame (aka Fares and other assorted revenue) 
Other_Funding_df <- clean_funding_df(7, "Other_Funding") %>% 
  group_by(UZA, Year) %>% 
  mutate(Other_Funding = sum(Other_Funding, na.rm = TRUE)) %>%  
  select(UZA, Year, Other_Funding) %>% 
  unique()

```

### Join FTA Funding Data frames
Joins the five above funding data frames into one comprehensive funding data frame. 
```{r, messages=FALSE}

# join all Funding dataframes
fta_funding_df <- full_join(Total_Funding_df, Federal_Funding_df, by = c("UZA", "Year")) %>% 
  full_join(State_Funding_df, by = c("UZA", "Year")) %>% 
  full_join(Local_Funding_df, by = c("UZA", "Year")) %>% 
  full_join(Other_Funding_df, by = c("UZA", "Year")) 

```

### Add Identifiers to FTA Data frame Through Relationship File and Join FTA data frames
This chunk reads in a file that contains the relationship between urbanized areas (used in our transit data) and metropolitan statistical areas (used in our quality of life data). It then tidies our final transit data frame. 
```{r, messages=FALSE}

fta_df <- left_join(primary_variable_fta_df, 
  final_transit_stations_df, by = c("UZA", "modes_clean"))%>%
  left_join(fta_funding_df, by=c("UZA", "Year")) %>% 
  unique()

# get census key dataframe
suppressMessages(
census_UZA_MSA_key <- read_csv("../census-FTA-joining-key/UZA_MSA_join.txt") %>% 
  mutate("GEO.id2" = CBSA) %>% 
  arrange(desc(UAPOP)) %>% 
  select(GEO.id2, UA, UANAME, UAPOP))

# join FTA dataframes here
final_fta_df <- left_join(census_UZA_MSA_key, fta_df,
                    by = c("UAPOP" = "UZA Population")) %>% 
  mutate(Year = as.numeric(Year))%>% 
  filter(Year > 2006, Year < 2018) %>% 
  filter(!is.na(`UZA Name`)) %>% 
  filter(UANAME != "Not in a 2010 urban area" ) %>% 
  filter(UAPOP>100000) %>% 
  select(-UANAME, -UZA)
  
```

## Create Final Data Frame
### Join Quality of Life and Transit Data
This chunk joins our quality of life and transit data frames by geographic ID and year. It also renames each of the columns for clarity. 
```{r, messages=FALSE}

# join final quality of life and fta datasets 
# and filter out non-urban areas and metro areas below 1 million population 
final_df <- left_join(qol_df, final_fta_df, by=c("GEO.id2", "Year"))%>% 
  filter(UA != 99999) %>% 
  mutate(gdp = gdp*1000000) %>% 
  select(-CENTLON, -CENTLAT, -UAPOP) %>% 
  filter(PopEstimate>=1000000)
 
# figure out which columns we want and rename columns accordingly
# clarify variables names
colnames(final_df)[1:29] <- c("msa_id",
                             "msa_name",
                             "pop_estimate_msa",
                             "year",
                             "median_household_income_msa",
                             "percent_commuting_msa",
                             "percent_unemployed_msa",
                             "percent_no_insurance_msa",
                             "percent_below_poverty_level",
                             "gdp_msa",
                             "intptlat",
                             "intptlon",
                             "ua_census_id",
                             "ua_sq_miles_2010",
                             "ua_fta_name",
                             "transit_modes",
                             "total_transit_expenses",
                             "total_fares",
                             "directional_route_miles",
                             "vehicle_hours",
                             "vehicle_miles",
                             "passenger_miles",
                             "passenger_trips",
                             "total_stations_2017",
                             "total_funding",
                             "federal_funding",
                             "state_funding",
                             "local_funding",
                             "other_funding")

```

### Adding Per Capita Variables 
This chunk adds new aggregate transit variables based on existing transit variables and adds them to our final data frame. 
```{r, messages = FALSE}

#Per Capita Transit Variables
final_df <- final_df %>% 
  mutate(per_capita_gdp = gdp_msa / pop_estimate_msa, 
         pmt_per_vrm = passenger_miles / vehicle_miles,
         pmt_per_vrh = passenger_miles / vehicle_hours,
         upt_per_vrh = passenger_trips / vehicle_hours,
         per_capita_vrm = vehicle_miles / pop_estimate_msa,
         per_capita_vrh = vehicle_hours / pop_estimate_msa,
         per_capita_pmt = passenger_miles / pop_estimate_msa,
         per_capita_upt = passenger_trips / pop_estimate_msa,
         recovery_ratio = total_fares / total_transit_expenses,
         fares_per_upt = total_fares / passenger_trips,
         cost_per_hour = total_transit_expenses / vehicle_hours,
         cost_per_trip = total_transit_expenses / passenger_trips,
         cost_per_pmt = total_transit_expenses / passenger_miles)
  
```

### Save Final Data Frame 
This converts our final data frame into a CSV as well as saving it as an RDS. This set of data is what the `transit_qol_df` in `gluskr` uses. 
```{r, messages = FALSE}

write.csv(final_df, "FINAL_DATA.csv", row.names=F)

saveRDS(final_df, file = "final_data.RDA")

```

# Visualizations and Modeling
## Initial ggplots
These are two examples of ggplots that we created. We created 20 ggplots with different transit variables as explanatory variables. These two ggplots, in which `percent_commuting_msa` and `per_capita_pmt` were utilized, had the highest R2 outputs.    
```{r, messages=FALSE}

## linear models with per_capita_gdp as response variable and different transit variables as explanatory variables

# explanatory variable: percent_commuting_msa
model_percent_commuting_msa <- lm(formula = per_capita_gdp ~ percent_commuting_msa, 
                                  data = final_df)

ggplot(final_df, aes(x = percent_commuting_msa, y = per_capita_gdp)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F)

summary(model_percent_commuting_msa)$adj.r.squared
summary(model_percent_commuting_msa)$r.squared

# explanatory variable: per_capita_pmt
model_per_capita_pmt <- lm(formula = per_capita_gdp ~ per_capita_pmt,
                           data = final_df)

ggplot(final_df, aes(x = per_capita_pmt, y = per_capita_gdp)) + 
  geom_point() +
  geom_smooth(method = 'lm', se = F)

summary(model_per_capita_pmt)$adj.r.squared
summary(model_per_capita_pmt)$r.squared

```

## Modeling
This function take in up to five variables and uses them as explanatory variables in a regression tree with per capita GDP as the response variable. To create the regression tree we first selected out all columns in the final data frame that weren't being used in this model. We then split up our data 80-20 into training and testing subsets. We then were able to get the resulting MAE, MSE, and R2 from our regression tree. This function returns a data frame with the resulting MAE, MSE, and R2 of the regression tree model for the given explanatory variables.
```{r, messages=FALSE}

# function to make a regression tree model with response variable per_capita_gdp
gdp_model <- function(explanatory_variables_eq, input1, input2=NULL, input3=NULL, input4=NULL, input5=NULL) {
  
  set.seed(2)

  select1 <- substitute(input1)
  select2 <- substitute(input2)
  select3 <- substitute(input3)
  select4 <- substitute(input4)
  select5 <- substitute(input5)
  
  if(is.null(input2)){
    cleaned_data <- final_df %>% 
    select(per_capita_gdp, select1)
  }else if(is.null(input3)){
    cleaned_data <- final_df %>% 
    select(per_capita_gdp, select1, select2)
  }else if(is.null(input4)){
    cleaned_data <- final_df %>% 
    select(per_capita_gdp, select1, select2, select3)
  }else if(is.null(input5)){
    cleaned_data <- final_df %>% 
    select(per_capita_gdp, select1, select2, select3, select4)
  }else{
    cleaned_data <- final_df %>% 
    select(per_capita_gdp, select1, select2, select3, select4, select5)}
  
  # only take rows with all data present
  final_cleaned_data <- cleaned_data[complete.cases(cleaned_data),]
  
  n <- nrow(final_cleaned_data)
  
  # select approx 80% of the row numbers between 1 and n
  train_id <- sample(1:n, size=round(n*0.8)) 

  # the data we'll train the model on
  train_data <- final_cleaned_data[train_id,]

  # the data we'll test the model on 
  test_data <- final_cleaned_data[-train_id,]
  
  # final_train_data <- train_data %>% drop_na("per_capita_gdp")
  
  regression_formula <- paste0("per_capita_gdp ~ ", explanatory_variables_eq)
  
  regression_tree <- rpart(regression_formula, data=train_data)
  predictions <- as.vector(predict(regression_tree, test_data)) # a vector of probabilities
  final_test_data <- test_data %>% mutate(prediction = predictions)
  final_test_data$truth <- final_test_data[,"per_capita_gdp"]
  results <- final_test_data %>%
    summarise("Explanatory Variables" = explanatory_variables_eq,
              #Percent_Rows = round(n/3416, 2),
              MSE = formatC( (1/n() * sum((truth - predictions)^2, na.rm = T)), format = "e", digits = 2),
              MAE  = round(1/n() * sum(abs(truth - predictions), na.rm = T), 0),
              R2   = round(cor(truth, predictions),3))
  
  return(results)
}


```

## Begin Modeling
Because our ggplots (of transit variables vs. per capita GDP) had very weak R2 values, we did further exploratory modeling using the above function to determine the best transit variables to predict per capita GDP. 
```{r, messages=FALSE}

# use above function on each transit variable as an explanatory variable
a <- gdp_model("percent_commuting_msa", "percent_commuting_msa")
b <- gdp_model("total_transit_expenses", "total_transit_expenses")
c <- gdp_model("directional_route_miles", "directional_route_miles")
d <- gdp_model("vehicle_hours", "vehicle_hours")
e <- gdp_model("passenger_miles", "passenger_miles")
f <- gdp_model("passenger_trips", "passenger_trips")
g <- gdp_model("total_funding", "total_funding")
h <- gdp_model("pmt_per_vrm", "pmt_per_vrm")
i <- gdp_model("pmt_per_vrh", "pmt_per_vrh")
j <- gdp_model("upt_per_vrh", "upt_per_vrh")
k <- gdp_model("per_capita_vrm", "per_capita_vrm")
l <- gdp_model("per_capita_vrh", "per_capita_vrh")
m <- gdp_model("per_capita_pmt", "per_capita_pmt")
n <- gdp_model("per_capita_upt", "per_capita_upt")
o <- gdp_model("recovery_ratio", "recovery_ratio")
p <- gdp_model("fares_per_upt", "fares_per_upt")
q <- gdp_model("cost_per_hour", "cost_per_hour")
r <- gdp_model("cost_per_trip", "cost_per_trip")
s <- gdp_model("cost_per_pmt", "cost_per_pmt")

model_results <- rbind(a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s)

kable(arrange(model_results, desc(as.numeric(R2))), 
      caption = "Exploratory Models")

```

## Refined Modeling
Based on our above modeling we determined that we should do further modeling on `percent_commuting_msa`, `per_capita_vrm`, `per_capita_upt`, `per_capita_vrh` and `per_capita_pmt`. In this code chunk we use our `gdp_model` function with every possible combination of these five explanatory variables and return the results in a data frame with each model's resulting MAE, MSE, and R2.s
```{r, messages=FALSE}

# get model results 

# one explanatory variable
model0 <- gdp_model("per_capita_vrm", "per_capita_vrm")
model1 <- gdp_model("percent_commuting_msa", "percent_commuting_msa")
model2 <- gdp_model("per_capita_pmt","per_capita_pmt")
model3 <- gdp_model("per_capita_vrh", "per_capita_vrh")
model4 <- gdp_model("per_capita_upt", "per_capita_upt")

# two explanatory variables
model5 <- gdp_model("per_capita_vrm + percent_commuting_msa", 
                    "per_capita_vrm", "percent_commuting_msa")
model6 <- gdp_model("per_capita_vrm + per_capita_pmt", 
                    "per_capita_vrm", "per_capita_pmt")
model7 <- gdp_model("per_capita_vrm + per_capita_vrh", 
                    "per_capita_vrm", "per_capita_vrh")
model8 <- gdp_model("per_capita_vrm + per_capita_upt", 
                    "per_capita_vrm", "per_capita_upt")
model9 <- gdp_model("percent_commuting_msa + per_capita_pmt",
                    "percent_commuting_msa", "per_capita_pmt")
model10 <- gdp_model("percent_commuting_msa + per_capita_vrh", 
                    "percent_commuting_msa", "per_capita_vrh")
model11 <- gdp_model("percent_commuting_msa + per_capita_upt", 
                    "percent_commuting_msa", "per_capita_upt")
model12 <- gdp_model("per_capita_upt + per_capita_vrh", 
                    "per_capita_upt", "per_capita_vrh")
model13 <- gdp_model("per_capita_upt + per_capita_pmt", 
                    "per_capita_upt", "per_capita_pmt")
model14 <- gdp_model("per_capita_vrh + per_capita_pmt", 
                    "per_capita_vrh", "per_capita_pmt")

# three explanatory variables
model15 <- gdp_model("per_capita_vrm + per_capita_pmt + percent_commuting_msa", 
                     "per_capita_vrm", "per_capita_pmt", "percent_commuting_msa")
model16 <- gdp_model("per_capita_vrm + per_capita_pmt + per_capita_vrh", 
                     "per_capita_vrm","per_capita_pmt", "per_capita_vrh")
model17 <- gdp_model("per_capita_vrm + per_capita_pmt + per_capita_upt", 
                     "per_capita_vrm","per_capita_pmt", "per_capita_upt")
model18 <- gdp_model("per_capita_vrm + percent_commuting_msa + per_capita_vrh",
                     "per_capita_vrm", "percent_commuting_msa", "per_capita_vrh")
model19 <- gdp_model("per_capita_vrm + percent_commuting_msa + per_capita_upt",
                     "per_capita_vrm", "percent_commuting_msa", "per_capita_upt")
model20 <- gdp_model("per_capita_vrm + per_capita_upt + per_capita_vrh",
                     "per_capita_vrm", "per_capita_upt", "per_capita_vrh")
model21 <- gdp_model("percent_commuting_msa + per_capita_upt + per_capita_vrh",
                     "percent_commuting_msa", "per_capita_upt", "per_capita_vrh")
model22 <- gdp_model("percent_commuting_msa + per_capita_upt + per_capita_pmt",
                     "percent_commuting_msa", "per_capita_upt", "per_capita_pmt")
model23 <- gdp_model("percent_commuting_msa + per_capita_vrh + per_capita_pmt",
                     "percent_commuting_msa", "per_capita_vrh", "per_capita_pmt")
model24 <- gdp_model("per_capita_vrh + per_capita_upt + per_capita_pmt",
                     "per_capita_vrh", "per_capita_upt", "per_capita_pmt")

# four explanatory variables
model25 <- gdp_model(
  "per_capita_vrm + percent_commuting_msa + per_capita_pmt + per_capita_vrh", 
  "per_capita_vrm", "percent_commuting_msa",
  "per_capita_pmt","per_capita_vrh")
model26 <- gdp_model(
  "per_capita_upt + percent_commuting_msa + per_capita_pmt + per_capita_vrh", 
  "per_capita_upt", "percent_commuting_msa",
  "per_capita_pmt","per_capita_vrh")
model27 <- gdp_model(
  "per_capita_upt + per_capita_vrm + per_capita_pmt + per_capita_vrh", 
  "per_capita_upt", "per_capita_vrm",
  "per_capita_pmt","per_capita_vrh")
model28 <- gdp_model(
  "per_capita_upt + per_capita_vrm + percent_commuting_msa + per_capita_vrh", 
  "per_capita_upt", "per_capita_vrm",
  "percent_commuting_msa","per_capita_vrh")
model29 <-  gdp_model(
  "per_capita_upt + per_capita_vrm + percent_commuting_msa + per_capita_pmt", 
  "per_capita_upt", "per_capita_vrm",
  "percent_commuting_msa","per_capita_pmt")

# all five explanatory variables
model30 <-  gdp_model(
  "per_capita_vrm + percent_commuting_msa + per_capita_pmt + per_capita_vrh + per_capita_upt", 
  "per_capita_vrm", "percent_commuting_msa",
  "per_capita_pmt","per_capita_vrh","per_capita_upt")

final_model <- rbind(model0, model1, model2, model3, model4, model5, 
                     model6, model7, model8, model9, model10, 
                     model11, model12, model13, model14, model15,
                     model16, model17, model18, model19, model20,
                     model21, model22, model23, model24, model25, 
                     model26, model27, model28, model29, model30)

kable(arrange(final_model, as.numeric(MAE)), caption="Final Models, Arranged by MAE")

```

## Results:

## Diagnostics:

## Conclusion: